EGGs are adhesive!\\ An algebraic account of equality saturation

### 2025-09-25 jlamp/jlamp.tex
minori
minima

Building on the criterion developed in~\cite{CastelnovoGM24}, this work proves that both hypergraphs with equivalence
and EGGs form an 
$\mathcal{M}$-adhesive category for a suitable choice of $\mathcal{M}$.
The advantages from this characterisation are two-fold. On the one side, 
we put the benefits \emph{per se} of a formal presentation, making precise the properties of the data structure. 
On the other side, describing the optimisation steps via the DPO approach
offers the tools for modelling  parallel and concurrent execution
and for proving  confluence and termination.

\emph{Synopsis}
The paper has the following structure. 
In Section~\ref{sec:ade} we briefly recall 
the theory of $\mathcal{M}$-adhesive categories
and of kernel pairs.
In Section~\ref{sec:hyper} we present the graphical structures of our interest, 
 (labelled) hypergraphs and term graphs, and we provide a
functorial characterisation, which allows for proving their adhesivity properties.
This is expanded in Section~\ref{hypereq} for proving the $\mathcal{M}$-adhesivity
of hypergraphs 
and term graphs with equivalence and in Section~\ref{eggs} of
their variants where equivalences are closed with respect to operator application,
thus subsuming EGGs.
%
In Section~\ref{rewriting} we put the machinery at work, showing how the optimisation steps
can be rephrased as the application of term graph rewriting rules.
%
Finally, in Section~\ref{conclusioni} we draw our conclusions, hint at future endeavours and offer some 
brief remarks on related works.
%
For the sake of space, the proofs appear in the appendices. 
%For the sake of completeness, 
%and in order to fix the notation, we prove all the results 
%recalled in the background section, besides those that are original 
%to our work.

@@

		      \node [below=.5cm, align=flush center,text width=2cm] at (-1.15,-0.5)
		{
			Figure $(2)$.
		};
	\end{pgfonlayer}
	\begin{pgfonlayer}{edgelayer}
		\draw(a1att.center) to (v1);
		\draw(a2att.center) to (v2);
		\draw[out=0, in=-120] (w1) to (divfst.center);
		\draw[out=0, in=120](w2) to (divsnd.center);
		\draw(divout.center) to (w);
	\end{pgfonlayer}
	\begin{pgfonlayer}{eqlayer}
		\draw[dashed, rounded corners] (-1.7, -0.8) rectangle (-0.6, 0.8);
	\end{pgfonlayer}
\end{tikzpicture}
\quad 
\begin{tikzpicture}[baseline=(w)]
	\begin{pgfonlayer}{nodelayer}
		\node[style=small box](a) at (-1.8, 0){$a$};
		\node[style=none] (aatt) at (-1.5, 0){};
		\node[style=node](v) at (-0.9, 0){};
		\node[style=medium box](div) at (0, 0){$/$};
		\node[style=none](divfst) at (-0.3, -0.3){};
		\node[style=none] (divsnd) at (-0.3, 0.3){};
		\node[style=none] (divout) at (0.3, 0){};
		\node[style=node] (w) at (0.9, 0){};
		      \node [below=.5cm, align=flush center,text width=2cm] at (-0.9,-0.5)
		{
			Figure $(3)$.
		};
	\end{pgfonlayer}
	\begin{pgfonlayer}{edgelayer}
		\draw(aatt.center) to (v);
		\draw[out=-60, in=-180] (v) to (divfst.center);
		\draw[out=60, in=-180](v) to (divsnd.center);
		\draw(divout.center) to (w);
	\end{pgfonlayer}
\end{tikzpicture}
\quad 
\begin{tikzpicture}[baseline=(w)]
	\begin{pgfonlayer}{nodelayer}
		\node[style=node](w1) at (-0.9, -0.5){};
		\node[style=small box](a2) at (-2.3, 0){$a$};
		\node[style=none] (a2att) at (-2.3, 0){};
		\node[style=node](v) at (-1.4, 0){};
		\node[style=node](w2) at (-0.9, 0.5){};
		\node[style=medium box](div) at (0, 0){$/$};
		\node[style=none](divfst) at (-0.3, -0.3){};
		\node[style=none] (divsnd) at (-0.3, 0.3){};
		\node[style=none] (divout) at (0.3, 0){};
		\node[style=node] (w) at (0.9, 0){};

@@

The fact that the right-hand side is a regular mono ensures that if the e-hypergraph to whom a rule is applied is acyclic,
the same property holds for the e-hypergraph obtained as the result.
%
Note however that cycles may implicitly occur due to the presence of the equivalence 
on nodes for what are called collapsing rules, i.e. where the right-hand side is a single variable. 
%
Looking again at the key example of~\cite{WillseyNWFTP21}, consider the rule $1 \times x = x$,
which can be modelled as the DPO rule below
\[
\hspace{.25cm}
\xymatrix{        
	\begin{tikzpicture}[baseline=(w.base)]\begin{pgfonlayer}{nodelayer}
			\node[style=node](v1) at (-0.9, 0.3){};
			\node[style=node](v2) at (-0.9, -0.3){};
			\node[style=medium box] at (0, 0){$\times$};
			\node[style=none](divfst) at (-0.3, 0.3){};
			\node[style=none](divsnd) at (-0.3, -0.3){};
			\node[style=none](divout) at (0.3, 0){};
			\node[style=node] (w) at (0.9, 0){};
			\node[style=small box] at (-2.3, 0.3){$1$};
			\node[style=none](one) at (-2, 0.3){};
			\node[style=node](z) at (-1.4, 0.3){};
		\end{pgfonlayer}        
		\begin{pgfonlayer}{edgelayer}
			\draw(v1) to (divfst.center);
			\draw(v2) to (divsnd.center);
			\draw(divout.center) to (w);
			\draw(one.center) to (z);
		\end{pgfonlayer}
		\begin{pgfonlayer}{eqlayer}
			\draw[dashed, rounded corners](-1.7, 0.6) rectangle (-0.6, -0);
		\end{pgfonlayer}\begin{pgfonlayer}{background}
			\draw[color=white] (-1.5, 0.2) rectangle (1.5, -0.2);
		\end{pgfonlayer}
	\end{tikzpicture}
	\ar@{=>}[r] &	\hspace{10pt}
	\begin{tikzpicture}[baseline=(w.base)]\begin{pgfonlayer}{nodelayer}
			\node[style=node](v1) at (-0.9, 0.3){};
			\node[style=node](v2) at (-0.9, -0.3){};
			\node[style=medium box] at (0, 0){$\times$};
			\node[style=none](divfst) at (-0.3, 0.3){};
			\node[style=none](divsnd) at (-0.3, -0.3){};
			\node[style=none](divout) at (0.3, 0){};
			\node[style=node] (w) at (0.9, 0){};
			\node[style=small box] at (-2.3, 0.3){$1$};
			\node[style=none](one) at (-2, 0.3){};
			\node[style=node](z) at (-1.4, 0.3){};
		\end{pgfonlayer}        
		\begin{pgfonlayer}{edgelayer}
			\draw(v1) to (divfst.center);
			\draw(v2) to (divsnd.center);
			\draw(divout.center) to (w);
			\draw(one.center) to (z);
		\end{pgfonlayer}
		\begin{pgfonlayer}{eqlayer}
			\draw[dashed, rounded corners](-1.7, 0.6) rectangle (-0.6, -0);
		\end{pgfonlayer}
		\begin{pgfonlayer}{background}
			\draw[color=white] (-1.5, 0.2) rectangle (1.5, -0.2);
		\end{pgfonlayer}
		\draw[dashed,rounded corners]
		([shift={(0,0.2)}]v2.north) --
		([shift={(-0.2,0)}]v2.west) --
		(-.9,-1.2)--(.9,-1.2)--([shift={(0.2,0)}]w.east)--
		([shift={(0,0.2)}]w.north)  -- ([shift={(-0.2,0)}]w.west)-- (0.7,-0.8)--(-0.7,-0.8)--([shift={(0.2,0)}]v2.east)-- cycle;
	\end{tikzpicture}
}
\]
%
Intuitively, the e-term graph obtained after the application of the rule seem to represent a recursive term such 
as $rec_x. 1 \times x$, highlighting the difficulty of extracting a term representation out of an e-term graph.

The fact that the left-hand side of a rule is the identity allows for the repeated application of such rule, hence the 
possibility to keep on performing the same rewriting step.
This is forbidden by using rules \emph{with negative application conditions}, given as the usual span 
$L \leftarrow K \rightarrow R$ plus an additional arrow $n: L\rightarrow N$, such that a match $m: L \to G$ is admissible if it cannot
be factorised through $n$~\cite{HabelHT96}. For e-term graphs and rules $L \leftarrow L \rightarrow R$, it suffices to choose $n$ 
as the right-hand side itself.
Indeed, consider again the rule $x / x \to 1$: once $1$ has been added once to the equivalence class of $x / x$, it is not necessary to

%
The theory of $\mathcal{M}$-adhesivity carries on for rules with negative application conditions, see~\cite{ehrig2012,ehrig2014adhesive}.

>>>>

Building on the criterion developed in~\cite{CastelnovoGM24}, this work considers a family of categories of hypergraphs
with equivalence, including EGGs, and prove that they form an 
$\mathcal{M}$-adhesive category for a suitable choice of $\mathcal{M}$.
The advantages from this characterisation are two-fold. On the one side, 
we put the benefits \emph{per se} of a formal presentation, making precise 
the properties of the data structures and the correspondence between them.
On the other side, describing the optimisation steps via the DPO approach
offers the tools for modelling  parallel and concurrent execution
and for proving  confluence and termination.

\emph{Synopsis}
The paper has the following structure. 
In Section~\ref{sec:ade} we briefly recall 
the theory of $\mathcal{M}$-adhesive categories
and of kernel pairs.
In Section~\ref{sec:hyper} we present the graphical structures of our interest, 
 (labelled) hypergraphs and term graphs, and we provide a
functorial characterisation, which allows for proving their adhesivity properties.
This is expanded in Section~\ref{hypereq} for proving the $\mathcal{M}$-adhesivity
of hypergraphs 
and term graphs with equivalence and in Section~\ref{eggs} of
a few variants where equivalences are closed with respect to operator application,
thus subsuming EGGs.
%
In Section~\ref{rewriting} we put the machinery at work, showing how the optimisation steps
can be rephrased as the application of DPO rewriting rules, at the same time
highlighting the issues linked to the format of such rules.
%
Finally, in Section~\ref{conclusioni} we draw our conclusions, hint at future endeavours and offer some 
brief remarks on related works.

For the sake of readability, all the proofs appear in the appendices. 
This work is an extended version of~\cite{uovadesive}.
With respect to the conference paper, the way term graphs
are described is completely changed: they are now presented 
as an instance of the newly introduced category of 
left-monogamous hypergraphs. Also the notion of separated 
hypergraphs is original, so that Sections 3.1.1 and 3.2.1 are largely new,
and consequently so Section 4.1.1, 4.2 and 5. 
%
Finally, also Section~6 on the applications of DPO rewriting has been rewritten 
and expanded, trying to highlight the nuances of its use.
%For the sake of completeness, 
%and in order to fix the notation, we prove all the results 
%recalled in the background section, besides those that are original 
%to our work.

@@

		      \node [below=.5cm, align=flush center,text width=2cm] at (-1.15,-0.5)
		{
			Figure $(2)$.
		};
	\end{pgfonlayer}
	\begin{pgfonlayer}{edgelayer}
		\draw(a1att.center) to (v1);
		\draw(a2att.center) to (v2);
		\draw[out=0, in=-120] (w1) to (divfst.center);
		\draw[out=0, in=120](w2) to (divsnd.center);
		\draw(divout.center) to (w);
	\end{pgfonlayer}
	\begin{pgfonlayer}{eqlayer}
		\draw[dashed, rounded corners] (-1.7, 0.2) rectangle (-0.6, 0.8);
		\draw[dashed, rounded corners] (-1.7, -0.8) rectangle (-0.6, -0.2);
	\end{pgfonlayer}
\end{tikzpicture}
\quad 
\begin{tikzpicture}[baseline=(w)]
	\begin{pgfonlayer}{nodelayer}
		\node[style=small box](a) at (-1.8, 0){$a$};
		\node[style=none] (aatt) at (-1.5, 0){};
		\node[style=node](v) at (-0.9, 0){};
		\node[style=medium box](div) at (0, 0){$/$};
		\node[style=none](divfst) at (-0.3, -0.3){};
		\node[style=none] (divsnd) at (-0.3, 0.3){};
		\node[style=none] (divout) at (0.3, 0){};
		\node[style=node] (w) at (0.9, 0){};
		      \node [below=.5cm, align=flush center,text width=2cm] at (-0.9,-0.5)
		{
			Figure $(3)$.
		};
	\end{pgfonlayer}
	\begin{pgfonlayer}{edgelayer}
		\draw(aatt.center) to (v);
		\draw[out=-60, in=-180] (v) to (divfst.center);
		\draw[out=60, in=-180](v) to (divsnd.center);
		\draw(divout.center) to (w);
	\end{pgfonlayer}
\end{tikzpicture}
\quad 
\begin{tikzpicture}[baseline=(w)]
	\begin{pgfonlayer}{nodelayer}
		\node[style=node](w1) at (-0.9, -0.5){};
		\node[style=small box](a2) at (-2.3, 0){$a$};
		\node[style=none] (a2att) at (-2.3, 0){};
		\node[style=node](v) at (-1.4, 0){};
		\node[style=node](w2) at (-0.9, 0.5){};
		\node[style=medium box](div) at (0, 0){$/$};
		\node[style=none](divfst) at (-0.3, -0.3){};
		\node[style=none] (divsnd) at (-0.3, 0.3){};
		\node[style=none] (divout) at (0.3, 0){};
		\node[style=node] (w) at (0.9, 0){};

@@

The fact that the right-hand side is a regular mono ensures that if the e-hypergraph to whom a rule is applied is acyclic,
the same property holds for the e-hypergraph obtained as the result.
%
Note however that cycles may implicitly occur due to the presence of the equivalence 
on nodes for what are called collapsing rules, i.e. where the right-hand side is a single variable. 
%
Looking again at the key example of~\cite{WillseyNWFTP21}, consider the rule $1 \times x = x$,
which can be modelled as the DPO rule below
\[
\hspace{.25cm}
\xymatrix{        
	\begin{tikzpicture}[baseline=(w.base)]\begin{pgfonlayer}{nodelayer}
			\node[style=node](v1) at (-0.9, 0.3){};
			\node[style=node](v2) at (-0.9, -0.3){};
			\node[style=medium box] at (0, 0){$\times$};
			\node[style=none](divfst) at (-0.3, 0.3){};
			\node[style=none](divsnd) at (-0.3, -0.3){};
			\node[style=none](divout) at (0.3, 0){};
			\node[style=node] (w) at (0.9, 0){};
			\node[style=small box] at (-2.3, 0.3){$1$};
			\node[style=none](one) at (-2, 0.3){};
			\node[style=node](z) at (-1.4, 0.3){};
		\end{pgfonlayer}        
		\begin{pgfonlayer}{edgelayer}
			\draw(v1) to (divfst.center);
			\draw(v2) to (divsnd.center);
			\draw(divout.center) to (w);
			\draw(one.center) to (z);
		\end{pgfonlayer}
		\begin{pgfonlayer}{eqlayer}
			\draw[dashed, rounded corners](-1.7, 0.6) rectangle (-0.6, -0);
		\end{pgfonlayer}\begin{pgfonlayer}{background}
			\draw[color=white] (-1.5, 0.2) rectangle (1.5, -0.2);
		\end{pgfonlayer}
	\end{tikzpicture}
	\ar@{=>}[r] &	\hspace{12pt}
	\begin{tikzpicture}[baseline=(w.base)]\begin{pgfonlayer}{nodelayer}
			\node[style=node](v1) at (-0.9, 0.3){};
			\node[style=node](v2) at (-0.9, -0.3){};
			\node[style=medium box] at (0, 0){$\times$};
			\node[style=none](divfst) at (-0.3, 0.3){};
			\node[style=none](divsnd) at (-0.3, -0.3){};
			\node[style=none](divout) at (0.3, 0){};
			\node[style=node] (w) at (0.9, 0){};
			\node[style=small box] at (-2.3, 0.3){$1$};
			\node[style=none](one) at (-2, 0.3){};
			\node[style=node](z) at (-1.4, 0.3){};
		\end{pgfonlayer}        
		\begin{pgfonlayer}{edgelayer}
			\draw(v1) to (divfst.center);
			\draw(v2) to (divsnd.center);
			\draw(divout.center) to (w);
			\draw(one.center) to (z);
		\end{pgfonlayer}
		\begin{pgfonlayer}{eqlayer}
			\draw[dashed, rounded corners](-1.7, 0.6) rectangle (-0.6, -0);
		\end{pgfonlayer}
		\begin{pgfonlayer}{background}
			\draw[color=white] (-1.5, 0.2) rectangle (1.5, -0.2);
		\end{pgfonlayer}
		\draw[dashed,rounded corners]
		([shift={(0,0.2)}]v2.north) --
		([shift={(-0.2,0)}]v2.west) --
		(-.9,-1.2)--(.9,-1.2)--([shift={(0.2,0)}]w.east)--
		([shift={(0,0.2)}]w.north)  -- ([shift={(-0.2,0)}]w.west)-- (0.7,-0.8)--(-0.7,-0.8)--([shift={(0.2,0)}]v2.east)-- cycle;
	\end{tikzpicture}
}
\]
%
Intuitively, the e-term graph obtained after the application of the rule seem to represent a recursive term such 
as $rec_x. 1 \times x$, highlighting the difficulty of extracting a term representation out of an e-term graph.

The fact that the left-hand side of a rule is the identity allows for the repeated application of such rule, hence the 
possibility to keep on performing the same rewriting step.
This is forbidden by using rules \emph{with negative application conditions}, given as the usual span 
$L \leftarrow K \rightarrow R$ plus an additional arrow $n: L\rightarrow N$, such that a match $m: L \to G$ is admissible if it cannot
be factorised through $n$~\cite{HabelHT96}. For e-term graphs and rules $L \leftarrow L \rightarrow R$, it suffices to choose $n$ 
as the right-hand side itself.
Indeed, consider again the rule $x / x \to 1$: once $1$ has been added once to the equivalence class of $x / x$, it is not necessary to
keep on adding it, since e.g. the two pictures below represent the same EGG
\[xxx\]
%
The theory of $\mathcal{M}$-adhesivity carries on for rules with negative application conditions, see~\cite{ehrig2012,ehrig2014adhesive}.
